\plaintitle

### 2017-02-21 paper/paper.tex
Paper: Better 'study' section. 'methods' almost ready

The use of computers and digital instruments is common in nowadays musical performances.
One significant different between traditional and digital musical instruments performances is that physical effort is usually involve in the former, but can be totally absent in the later.
This difference presents several issues for live computer music.
For example, it might prevent the audience from understanding the performance.
On the extreme, the listener might question the "liveness" of a performance altogether when he can't tell the difference between live and pre-composed playback.
Besides, visible effort is an important expressive tool in many musical contexts.
TODO BETTER EXAMPLES
Consider, for example, a trumpet player and his outcurved veins in the neck while he reaches a high note, or a drummer, sweating during intense performance.

@@

In an initial study I examined to what extent physical effort can enhance expressivity in playing digital instruments.
TODO X participants were invited to play with the Schleikess based synthesiser and discuss their subjective experience.
Apart from playing, they could modify the amount of physical effort required by the instrument by changing the sensitivity of the interface to force.
Two sliders on the computer screen controlled the sensitivity, one for the pitch and one for the amplitude.
This way, when the sensitivity is low, a major physical force is needed to raise the pitch or amplitude, but when the sensitivity was high even minor changes affect the output significantly.
The participants were instructed that they need to stretch the elastic bands to play the instrument and that the sliders on the computer screen changes the sensitivity of the instrument.
However, the relationship between stretching the bands and the resulted sound was not explaind.

The participants were asked to familiarize themeselves with the instrument, understand the mapping between gestures and sound, and then set the sliders to the values they find most expressive.
Later, I discussed with each participant about the experience of playing the instrument.
The discussion concentrated around questions of how expressive and musical the instrument felt with differente sensitivity settings.

This type of exploratory initial study is designed to get basic insights about the relationships between physical effort and expressivity.
Provided that there is almost no empirical research in this field, this type of broad exploration is necessary to gain first insights.

@@

>>>>

The use of computers and digital instruments is common in nowadays musical performances.
TODO one more sentence here.
One significant difference between traditional and digital musical instruments performances is that physical effort is usually involve in the former, but can be totally absent in the later.
This difference presents several issues, most of these thorouly explored in the context of laptop music.

TODO paragraph explaining laptop music.

For example, a common criticism is that the audiance might fail to understand the performance.
On the extreme, a listener might question the "liveness" of a performance altogether when unable to tell the difference between live and pre-composed playback.
Besides, visible effort is an important expressive tool in many musical contexts.
TODO BETTER EXAMPLES
Consider, for example, a trumpet player and his outcurved veins in the neck while he reaches a high note, or a drummer, sweating during intense performance.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.9\columnwidth]{figures/pd_interface}
  \caption{The Schleikess graphical interface. Three sliders controll the sensitivity of the pitch band to force, the sensitivity of the amplitude band to force, and the overall volume of the instrument.}~\label{fig:pd-interface}
\end{figure}

In this initial study I examined to what extent physical effort can enhance expressivity in playing digital instruments.
TODO more specifically... Controllability, range of possibilities. Or maybe just in the discussion.

TODO X participants were invited to play with the Schleikess based synthesiser and discuss their subjective experience.
Apart from the physical interface, figure \ref{fig:pd-interface} shows the interface on the computer screen that was exposed to the participants.
The interface presents three sliders.
The first two controll the sensitivity of each elastic band to force.
So, when the sensitivity is set to low value, a major physical force is needed to raise the pitch or amplitude, but when the sensitivity is set to high value even minor changes affect the output significantly.
The last slider controlls the overall volume of the instrument and allows the participant to set it to a comfortable hearing level.

First, I explained to the participants that stretching the elastic bands controlls the instrument.
The relationship between stretching the bands and the resulted sound was not explaind.
Than, the participants were asked to play the instrument until they understand the mapping between gestures and sound.
During this time they could also set the volume to a comfortable hearing level.
After familiarize themselves with the instrument they were asked to set the sensitivity sliders to the values they like the most.
Later, I discussed the experience of playing the instrument with each participant.
The discussion concentrated around the question of why they choose these particular sensitivity settings.
In some of the discussions the participants referred to how expressive and controllable the instrument felt.
In these cases I asked them to elaborate and explain the terms, and why they felt this way.
However, participants that didn't mentioned these concepts in the first place were not forced to talk about them.

This type of exploratory initial study is designed to get basic insights about the relationships between physical effort and expressivity.
Provided that there is almost no empirical research in this field, this type of broad exploration is necessary to gain first insights, and to be able to design a more thorough experiment, as I will describe later.

@@

Most of the participants in the study, TODO out of TODO, understood how the instrument works by trying it out.
I explained the relationship between pulling the elasic bands and the resulted sound to those who find it hard to understand.
After the explaination they continued with the experiment like the rest of the participants.

TODO participants explicitly discussed the balance between having range of possibilities and controllability with relationship to the sensitivity of the instrument to force.
This was significantly apparant in the discussion of the pitch sensitivity.
TODO participants wanted to maximize the pitch range that the instrument produces by raising the sensitivity, but no participant set the pitch sensitivity to the maximum value.
TODO participants explicitly explained that they kept the pitch sensitivity below the maximum value to make sure they can control the instrument.
On the other hand, only TODO participants kept the pitch sensitivity low, and explained this preference by not liking notes with high pitch in general.

Most of the participants presented different explanations for choosing the sensitivity values for the pitch and for the amplitude.
Only TODO of the participants used the same reasoning for setting the two sliders.

TODO participants mentioned that playing the instrument felt like a physical exercise, is somewhat tedious, or made them pain over time.
Therefore, TODO prefered to minimize effort or movement while playing.
One participant, attempting to minimize the effort involve in large hand movement, described a completely different playing technique.
Instead of pull and release the bands, the participant held them in constant force and change the tention by pushing the pinky against the bands.
